---
title: "Outline"
output: html_document
---

```{r}
# These are ideas for headings to help us structure our work. Feel free to edit
```

1. Introduction: Who Are Data Scientists in Education? 

```{r}
# This where we can discuss the intended audience
```

2. How to Use This Book 

    - Suggested learning process: reading, doing the examples, applying to your own datasets  
    - How to submit datasets for practice  
    - Mention of why we're using the Tidyverse
    - Mention of the benefits of keeping R up-to-date, , learning to deal with functions that are no longer supported, and proactively writing code so it is easier to update in response to new versions

3. Data Science in Education

4. Unique Challenges

5. Foundational Skills  

    - Processing / preparing data
    - Plotting data
    - Importing / exporting data from various sources
    - Modeling data
    - Communicating / sharing results

6. Education Dataset Analysis Pipeline: Walk Through #1  

- Online Science Ed

7. Education Dataset Analysis Pipeline: Walk Through #2 

- Gradebook

8. Education Dataset Analysis Pipeline: Walk Through #3 

- Machine Learning

9. Education Dataset Analysis Pipeline: Walk Through #4  

- Networks

    - Add a discussion of use cases for network analysis to our existing social media example

10. Education Dataset Analysis Pipeline: Walk Through #5

- Aggregate Data

11. Education Dataset Analysis Pipeline: Walk Through #6

- Longitudinal Data

12. Advanced Uses

- Multi-Level Modelling

13. Applying New Skills in Workplace

14. Students Doing Data Science

15. Resources

16. Learning More
