---
title: "Outline"
output: html_document
---

```{r}
# These are ideas for headings to help us structure our work. Feel free to edit
```

1. Introduction: Who Are Data Scientists in Education? 

```{r}
# This where we can discuss the intended audience
```

2. How to Use This Book 

    - Suggested learning process: reading, doing the examples, applying to your own datasets  
    - How to submit datasets for practice  

4. The Role of the Data Scientist in Education  

3. Unique Challenges in Education Data 

5. Foundational Skills  

    - Processing / preparing data
    - Plotting data
    - Importing / exporting data from various sources
    - Modeling data
    - Communicating / sharing results

6. Advanced Uses  

    - Multi-level models 
    - Text analysis
    - Longitudinal analysis  

7. Education Dataset Analysis Pipeline: Walk Through #1  

8. Education Dataset Analysis Pipeline: Walk Through #2 

9. Education Dataset Analysis Pipeline: Walk Through #3 

10. Solutions for Adopting Data Science Techniques in Education 

11. Students' Use of Data Science Techniques

12. Resources

