---
title: 'Education Dataset Analysis Pipeline: Walkthrough #2'
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
```


## Introduction
The entry bar into data science in education might seem daunting to the typical K-12 classroom educator. The knowledge, concepts, and skills taught in this book are not typically taught to pre-service and in-service teachers alike.

Learning R helps K-12 teachers explore evidence of student learning, performance, and outcomes. Furthermore, R helps K-12 teachers to visually communicate this evidence to school community stakeholders including parents, students, school administrators, and others.

There are a variety of data sources to explore from the K-12 classroom. Student assessment scores is a typically cited example. However, more novel data sources are worth exploring as well. For example, the text from a teacher’s written classroom observation notes about a particular learner’s in-class behavior or emotional status can be analyzed for trends. Another example, is tapping into the exportable data available from common learning software or platforms popular in the K-12 education space.

This walkthrough goes through a series of analyses using the data science framework (link). The first analysis centers around a ubiquitous K-12 classroom tool: the gradebook. We use an Excel gradebook template [Assessment Types - Points](https://web.mit.edu/jabbott/www/excelgradetracker.html) and simulated student data. On your first read through of this section try using our simulated dataset found in this book's `data/` folder.

***

## Driving Question and Objectives

***

## Setting up RStudio
To get started first setup the RStudio environment by loading the libraries that will be used in this chapter. The *tidyverse* library should be familiar from previous chapters. The *readxl* library is used to read and import Excel spreadsheets since these file types are very common in K-12 education.

Make sure you have installed the packages in R on your computer before starting. If you have already installed the packages once before then skip this step.
```{r eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE}
# Install libraries
install.packages("tidyverse")
install.packages("readxl")
```

Next, load the libraries. These must be loaded each time we start a new project.
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Load libraries
library(tidyverse)
library(readxl)
```

***

## Import Data
Recall how the foundational skills section recommended favoring CSV files, or comma-separated-values files, when working with datasets in R. This is because CSV files, with the .csv file extension, are ubiquitous to the the digital world. However, it is understandable that data won't always come in the preferred file formats. Fortunately, R can import a variety of data file types. This walkthrough imports an Excel file because these file types, with the .xlsx extension, are more likely to be encountered in the K-12 education world than other typical file types that R users might encounter such as .csv, .txt, or .sav files.

This code uses the `read_excel()` function of the *readxl* package to find and read the data of the desired file. Note the file path that `read_excel()` takes to find the simulated dataset file named *ExcelGradeBook.xlsx* which sits in a folder on your computer if you have downloaded it. In the code below `ExcelGradeBook <- read_excel("path/to/file.xlsx", sheet = 1, skip = 10)` the part written `path/to/file.xlsx` is just pseudo code to remind you to swap in your own path to the file you want to import. Recall from the foundational skills section of this book that directories and file paths are important for finding files on your computer. The function `getwd()` will help locate your current working directory. This tells where on the computer R is currently working with files.

```{r eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE}
# See the current working directory
getwd()
```

For example, an R user on Linux or Mac might see their working directory as: `/home/username/Desktop`. A Windows user might see their working directory as: `C:\Users\Username\Desktop`.

From this location go deeper into files to find the desired file. For example, if you downloaded the book repository from Github to your Desktop the path to the Excel file might look like one of these below:

* `/home/username/Desktop/data-science-in-education/data/gradebooks/ExcelGradeBook.xlsx` (on Linux & Mac) 
* `C:\Users\Username\Desktop\data-science-in-education\data\gradebooks\ExcelGradeBook.xlsx` (on Windows)

After locating the sample Excel file run the code below to run the function `read_excel()` which reads and saves the data from *ExcelGradeBook.xlsx* to an object also called *ExcelGradeBook*. Note the two arguments specified in this code `sheet = 1` and `skip = 10`. This Excel file is similar to one you might encounter in real life with superfluous features that we are not interested in. This file has 3 different sheets and the first 10 rows contain things we won't need. Thus, `sheet = 1` tells `read_excel()` to just read the first sheet in the file and disregard the rest. While `skip = 10` tells `read_excel()` to skip reading the first 10 rows of the sheet and start reading from row 11 which is where the column headers and data actually start inside the Excel file. Remember to replace `path/to/file.xlsx` your own path to the file you want to import.

```{r eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE}
# Use readxl package to read and import file and assign it a name
ExcelGradeBook <- read_excel("path/to/file.xlsx", sheet = 1, skip = 10)
```

```{r eval = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
# Use readxl package to read and import file and assign it a name
ExcelGradeBook <- read_excel("./data/gradebooks/ExcelGradeBook.xlsx", sheet = 1, skip = 10)
```

A more beginner friendly way to import the dataset is to run this code to interactively select the file from a window.
```{r eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE}
# Opens a window to select a file using arguments
ExcelGradeBook <- read_excel(file.choose(), sheet = 1, skip = 10)
```

The *ExcelGradeBook* file has been imported into RStudio. Next, assign the data frame to a new name using the code below. Renaming cumbersome filenames can improve the readability of the code and make is easier for the user to call on the dataset later on in the code.
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Rename data frame
gradebook <- ExcelGradeBook
```

Your environment will now have two versions of the dataset. There is *ExcelGradeBook* which was the original dataset imported. There is also *gradebook* which is currently a copy of *ExcelGradeBook*. As you progress through this section, we will work primarily with the *gradebook* dataset. Additionally, while working onward in this section of the book, if you make a mistake and mess up the *gradebook* data frame and are not able to fix it, you can reset the data frame to return to the same state as the original *ExcelGradeBook* data frame by running `gradebook <- ExcelGradeBook` again. This will overwrite your messed up *gradebook* data frame with the originally imported *ExcelGradeBook* data frame. Afterwards, just continue running code from this point in the text.

***

## Tidy Data
This walkthrough uses an Excel data file because it is one that K-12 educators are likely to encounter. Moreover, the messy state of this file mirrors what might be encountered in real life. The Excel file contains more than one sheet, has rows we don't need, and uses column names that have spaces between words. All these things make the data tough to work with. The data is **not** tidy. These challenges could be fixed before importing the file into RStudio by deleting, for example, the unnecessary parts of the Excel file then saving it as a .csv file. However, the untidy Excel file provides realistic practice for tidying up the data programmatically using R itself.

First, modify the column names of the *gradebook* data frame to remove any spaces and replace them with an underscore. Using spaces as column names in R can present difficulties later on when working with the data.
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Rename columns to use underscore instead of spaces
names(gradebook) <- gsub(" ", "_", names(gradebook))
```

Review what the *gradebook* data frame looks like now. It shows 25 students and their individual values in various columns like *Projects* or *Formative_Assessments*.
```{r eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE}
view(gradebook)
```

The data frame looks cleaner now. However, there still are some things we can remove. For example, there are rows without any names in them. Also, there are entire columns that are unused and contain no data. These are called missing values and are denoted by *N/A*. Since our simulated classroom only has 25 learners and doesn't use all the columns for demographic information, we can safely remove these to tidy our dataset up even more.

This code removes the columns that contain no data.
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Removing columns with nothing but missing data
gradebook <- gradebook[colSums(!is.na(gradebook)) > 0]
```

Now that the extra rows have been removed, notice there are two columns, *Absent* and *Late*, that someone started putting data into but then decided to stop. Thus, these two columns didn't get removed by the last chunk of code because they technically contained some data in those columns. However, since the simulated teacher of this simulated class decided to abandon using the *Absent* and *Late* columns in this gradebook, we should remove it from our data frame as well.
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Remove a targeted column because we don't use Absent and Late at this school.
gradebook <- subset(gradebook, select = -c(Absent, Late))
```

Finally, remove the extra rows that have no data.
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Removing rows with nothing but missing data
gradebook <- na.omit(gradebook)
```

At last, the formerly untidy Excel sheet has been turned into a useful data frame. Inspect it once more to see the difference.
```{r eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE}
view(gradebook)
```

***

## Transform Data
R users transform data to facilitate working with the data during later phases of visualization and analysis. A few examples of data transformation include renaming variables, creating new variables, grouping data, and more. This code chunk first creates a new data frame named *classwork_df*, then selects particular variables from our gradebook data frame, and finally gathers all the homework data under new variables into new columns.
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Creates new data frame, selects desired variables from gradebook, and gathers all classwork scores into key/value pairs
classwork_df <- gradebook %>%
  select(Name, Running_Average, Letter_Grade, Homeworks, Classworks, Formative_Assessments, 
  Projects, Summative_Assessments, contains("Classwork_")) %>%
  gather(contains("Classwork_"), key = "Classwork_Number", value = "Score")
```

View the new data frame and note which variables were selected for this new data frame. Also, note how all the classwork scores were gathered under new variables. The `contains()` *function* searches for any column with the string *"Classwork_"*. The underscore makes sure the variables from *Classwork_1* all the way to *Classwork_17* are included. We will use this *classwork_df* data frame later.
```{r echo = TRUE, message = FALSE, warning = FALSE}
view(classwork_df)
```

***

## Visualize Data
Visual representations of data are more human friendly than just looking at numbers alone. For example, this next line of code shows a simple summary of the data by each column.
```{r eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE}
# Summary of the data by columns
summary(gradebook)
```

But R can do more than just print numbers to a screen. Use the ggplot function to graph some of the data to help get a better grasp of what the data looks like. This code uses ggplot to graph categorical variables into a bar graph. Here we can see the variable *Letter_Grade* is plotted on the x-axis showing the counts of each letter grade on the y-axis. 
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Bar graph for categorical variable
ggplot(data = gradebook) +
  geom_bar(mapping = aes(x = Letter_Grade, fill = Running_Average > 90)) +
  labs(title ="Bar Graph of Student Grades", x = "Letter Grades", y = "Count", fill = "A or Better")
```

Use the ggplot function to graph these continuous variables onto a scatterplot. The x-axis shows the learners' overall grade in percentages while the y-axis lists each individual learner. The argument `color = Running_Average > 90` colors the points to helps point out which learners have overall grades above 90%.
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Scatterplot of continuous variable
ggplot(data = gradebook) + 
  geom_point(mapping = aes(x = Running_Average, y = Name, color = Running_Average > 90)) +
  labs(title ="Scatterplot of Student Grades", x = "Overall Grade in Percentage", y = "Name", 
  color = "90% or Above")
```

We can layer different types of plots on top of each other. Here a scatterplot is layered with a regression line fitted to the data. The x-axis shows the homework grades while the y-axis shows the overall grades. The graph suggests a correlation between overall class grade and homework grades. As the homework grade goes up, the overall grade goes up too.
```{r echo = TRUE, message = FALSE, warning = FALSE}  
# Plot of covariance between homework score and grades by percent
ggplot(data = gradebook) + 
  geom_point(mapping = aes(x = Homeworks, y = Running_Average)) +
  geom_smooth(mapping = aes(x = Homeworks, y = Running_Average)) +
  labs(title ="Correlation Between Overall Grade and Homework", x = "Homework Grade in Percentage", 
  y = "Overall Grade in Percentage")
```

***

## Model Data

***

