# Walkthrough 1

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = F)
here::set_here(path = "~/documents/data-science-in-education")
```

# Background 

In the 2015-2016 and 2016-2017 school years, researchers at Michigan State University carried out a study on students' motivation to learn in online science classes. The online science classes were part of a statewide online course provider designed to supplement (and not replace) students' enrollment in their local school. For example, students may choose to enroll in an online physics class because one was not offered at their school (or they were not able to take it given their schedule).

The study involved a number of different data sources which were brought to bear to understand students' motivation:

1. A self-report survey for three distinct but related aspects of studnets' motivation
2. Log-trace data, such as data output from the learning management system
3. Achievement-related and gradebook data
4. Discussion board data
5. Achievement-related (i.e., final grade) data

First, these different data sources will be described in terms of how they were provided by the school.

## 1. Self-report survey 

This was data collected before the start of the course via self-report survey. The survey included 10 items, each corresponding to one of three *measures*, namely, for interest, utility value, and perceived competence:

1.	I think this course is an interesting subject. (Interest)
2.	What I am learning in this class is relevant to my life. (Utility value)
3.	I consider this topic to be one of my best subjects. (Perceive competence)
4.	I am not interested in this course. (Interest - reverse coded)
5.	I think I will like learning about this topic. (Interest)
6.	I think what we are studying in this course is useful for me to know. (Utility value)
7.	I don’t feel comfortable when it comes to answering questions in this area. (Perceived competence)
8.	I think this subject is interesting. (Interest)
9.	I find the content of this course to be personally meaningful. (Utility value)
10.	I’ve always wanted to learn more about this subject. (Interest)

## 2. Log-trace data 

Log-trace data is data generated from our interactions with digital technologies. In education, an increasingly common source of log-trace data is that generated from interactions with learning management systems. The data for this walk-through is a *summary of* log-trace data, namely, the number of minutes students spent on the course. Thus, while this data is rich, you can imagine even more complex sources of log-trace data (i.e. timestamps associated with when students started and stopped accessing the course!).

## 3. Achievement-related and gradebook data

This is a common source of data, namely, one associated with graded assignments students completed. In this walkthrough, we just examine students' final grade.

## 4. Discussion board data
<!-- NOTE - may not include this, as it is hard to confidently anonymize a medium-ish sized data set -->

Discussion board data is both rich and unstructured, in that it is primarily in the form of written text. We examine a small subset of the discussion board data in this walkthrough.

# Processing the data

```{r, loading-packages}
library(readxl)
library(tidyverse)
library(lubridate)
library(here)
```

```{r}
# Gradebook and log-trace data for F15 and S16 semesters
s12_course_data <- read_csv(here::here("data", "online-science-motivation", "raw", "s12-course-data.csv"))

# Pre-survey for the F15 and S16 semesters
s12_pre_survey  <- read_csv(here::here("data", "online-science-motivation", "raw", "s12-pre-survey.csv")) 

# Log-trace data for F15 and S16 semesters - ts is for time spent
s12_time_spent <- read_csv(here::here("data", "online-science-motivation", "raw", "s12-course-minutes.csv"))
```

## Viewing the data

```{r}
s12_pre_survey 
s12_course_data
s12_time_spent
```

## Joining the data

To join the course data and pre-survey data, we need to create similar *keys*.

Namely, both have variables for the course and the student, thought they have different names in each.

Let's start with the pre-survey data.

```{r}
s12_pre_survey <- s12_pre_survey %>% 
    rename(student_id = RespondentId,
           course_id = opdata_CourseID)

s12_pre_survey
```

Looks better now!

Let's proceed to the course data.

```{r}
s12_course_data <- s12_course_data %>% 
    rename(student_id = Bb_UserPK,
           course_id = CourseSectionOrigID)

s12_course_data
```

Now we can join these using the **dplyr** function, `left_join()`:

```{r}
dat <- left_join(s12_course_data, s12_pre_survey, by = c("student_id", "course_id"))

ddd
```

Note that they're now combined, even though the course data has many more rows: The pre_survey data has been join for each student by course combination.

