---
title: "Education Dataset Analysis Pipeline: Walkthrough #5"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = F)

library(tidyverse)
library(here)
```

## Introduction

A common situation encountered when using data for analyzing the education sector, particularly for analysts who are not directly working with schools or districts, is the prevalence of aggregate data. [Aggregate data refers to numerical or non-numerical information that is (1) collected from multiple sources and/or on multiple measures, variables, or individuals and (2) compiled into data summaries or summary reports, typically for the purposes of public reporting or statistical analysis.](https://www.edglossary.org/aggregate-data/) Example of publicly available aggregate data include school-level graduation rates or state test proficiency scores by grade and subject.

These datasets are large, lagging, and often suppressed to protect privacy. Because of their coarseness, they can be difficult to use in decision-making. However, these datasets are available to all analysts for gleaning insights about education and can be used for landscape analyses or to supplement other analyses.

## Data

There are many publicly available aggregate datasets. On the federal level, examples include:

- [EdFacts](https://www2.ed.gov/about/inits/ed/edfacts/data-files/index.html), which includes state assements and adjusted cohort graduation rates. These data are school- and district-level.

- [Civil Rights Data Collection (CRDC)](https://www2.ed.gov/about/offices/list/ocr/data.html), which reports many different variables on educational program and services disaggregated by race/ethnicity, sex, limited English proficiency, and disability. These data are school-level.

- [National Assessment for Educational Progress (NAEP) Data](https://nces.ed.gov/nationsreportcard/researchcenter/datatools.aspx), a nationally representative assessment of educational progress in the United States. These data are national-, state-, and select city-level.

### Import

For the purposes of this walkthrough, we will be looking at NAEP 2017 results. These files, pulled from the [Nation's Report Card](https://www.nationsreportcard.gov/math_2017/nation/achievement?grade=4), are woefully not tidy and instead come with a lot of formatting and merged cells. Therefore, wrangling the dataset is necessary before any analysis can occur.

```{r}
naep17 <-
    readxl::read_xlsx(here::here("data", "naep", "2017_Results_Appendix_Math_State.xlsx"), sheet = 5, skip = 8)
```

### Cleaning

Once imported, the dataset contains a column with different states/jurisdiction, columns with the various years and the proficiency rates for each one, and miscellaenous columns denoting statistic significance. There are also rows at the bottom with additional information on the dataset. For the purpose of our walkthrough, we want the dataset to be as clean and simple as possible. We will remove the columns with significance status, the rows with additional information, and transform the proficiency rates to numeric.

```{r}
head(naep17)
```

```{r}
naep17_clean <-
    naep17 %>% 
    select(-dplyr::contains("X")) %>%  # selecting only years allowing accomodations
    slice(-57:-62) %>% # remove rows containing additional information
    mutate_at(vars(2:13), funs(as.numeric(.)/100)) %>% # transforming proficiency rates to numeric decimals
    select(-2:-4) %>% # years allowing accomodations
    rename(state = `State/jurisdiction`,
           `2000` = `2000__1`) %>% 
    gather(key = "year", value = "proficiency", -1) # making a tidy dataset
```

## Analysis

### Discovering Distributions

### Understanding Trends

### Making Comparisons

### Finding Inequities

### Working with Suppression

### Statistical Inferences

## Results

## Conclusion

## Appendix

Loeb, S., Dynarski, S., McFarland, D., Morris, P., Reardon, S., & Reber, S. (2017). [Descriptive analysis in education: A guide for researchers.](https://ies.ed.gov/ncee/pubs/20174023/pdf/20174023.pdf) (NCEE 2017–4023). Washington, DC: U.S. Department
of Education, Institute of Education Sciences, National Center for Education Evaluation and Regional Assistance. 


