---
title: "Chapter 11 Walkthrough: Longitudinal Analysis"
author: 
date: 
output: html_document
---

# Vocabulary 

Read in
Aggregate data 
Vector 
List
Tidy format
Subset
`select_at`
`mutate`

# 11.1 Introduction 

The authors chose to use a publicly available dataset for this walkthrough. Like most publicly available datasets, this one contains aggregate data. 

# 11.2 Special Education Child Count and Environment Data

You can download these datasets on the United States Department of Education website: [https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/index.html#bccee]("https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/index.html#bccee").

# 11.2.1 Reading In One Dataset 

When you are designing an analysis that uses multiple datasets that all have the same structure, you can read in each dataset using one chunk of code. This chunk of code will store each dataset as an element of a list. 

Before doing that, you can explore just one of the datasets briefly to see what you can learn about the dataset's structure. Clues from this initial exploration of one dataset informs how you read in all the datasets at once later on. For example, this first dataset has some lines at the top that don't actually contain data: 

```{r}
library(tidyverse)

read_csv("10a-walkthrough-6-data/bchildcountandedenvironments2012.csv")
```

The rows containing "Extraction Date:", "Updated:" and "Revised:" aren't actually rows. They're notes the authors left at the top of the dataset to show when the dataset was changed. 

`read_csv` uses the first row as the variable names unless told otherwise, so we need to explicitly tell `read_csv` to skip those lines using the `skip` argument. If we don't, `read_csv` assumes the very first line--the one that says "Extraction Date:"--is the correct row of variable names. That's why calling `read_csv` without the `skip` argument results in a bunch of column names like `X4`. When there's no obvious column name to read in, `read_csv` names them `X[...]` and let's you know in a warning message. 

Try using `skip = 4` in your call to `read_csv`:

```{r}
read_csv("10a-walkthrough-6-data/bchildcountandedenvironments2012.csv", skip = 4)
```

The `skip` argument told `read_csv` to think of the line containing "Year", "State Name", and so on as the first line. That's why using `skip = 4` gave you a dataset with "Year", "State Name", and so on as variable names. 

# 11.2.1 Reading In Many Datasets 

Will the `read_csv` and `skip = 4` combination work on all our datasets? To find out, we'll use this strategy: 

 - Store a vector of filenames and paths in a list. These paths point to our datasets
 - Pass the list of filenames as arguments to `read_csv` using `purrr::map`, including `skip = 4` in our `read_csv` call
 - Examine the new list of datasets to see if the variable names are correct in each

Imagine a widget-making machine that works by acting on raw materials it receives on a conveyer belt. This widget-making machine follows one set of instructions on each of the raw materials it receives. You are the operator of the machine and you design instructions to get a widget out of the raw materials. Your plan might look something like this: 

 - *Raw materials*: a list of filenames and their paths 
 - *Widget-making machine*: `purrr:map()`
 - *Widget-making instructions*: `read_csv(path, skip = 4) 
 - *Expected widgets*: a list of datasets

Let's create the raw materials first. Use `list.files` to make a list of filenames and their paths and name that list `filenames`: 

```{r get filenames}
# Get filenames from the data folder 
filenames <- list.files("10a-walkthrough-6-data", full.names = TRUE)

# A list of filenames and paths
filenames
```

That made a list of six filenames, one for each year of child count data stored in the data folder. Now pass our raw materials, the list called `filenames`, to our widget-making machine called `map` and give the machine the instructions `read_csv(., skip = 4)`. Name the list of widgets it cranks out `all_files`:

```{r read list of CSVs}
# Pass filenames to map and read_csv
all_files <- filenames %>% map(., ~read_csv(., skip = 4))

# A list of datasets
all_files
```

It is important here to think ahead. Ultimately we want to combine all the datasets in `all_files` into one dataset using `bind_rows`. But that will only work if all the datasets in our list have the same number of columns and the same column names. We can check our column names by using `map` and `names`: 

```{r include=FALSE}
all_files %>% map(names)
```

We can use `identical` to test the variables of two datasets. For example, we can show the variable names of the first and datasets don't match, but the names from the third and second do match. 

```{r check names}
# Variables of first and second dataset don't match
identical(names(all_files[[1]]), names(all_files[[2]]))

# Variables of third and second files match
identical(names(all_files[[3]]), names(all_files[[2]]))
```

And we can check the number of columns by using `map` and `ncol`:

```{r}
all_files %>% map(ncol)
```

If this is your first time working on education data, congratulations on finding an extremely common problem! By now you've not only noticed that the number of columns don't match across datasets, but the column names don't match either. This is a problem because we won't be able to combine the datasets into one long one. When we try, `bind_rows` returns a dataets with 100 columns instead of the expected 50. 

```{r combine datasets}
bind_rows(all_files)
```

We'll correct this in the next section by selecting and renaming our variables, but it's good to notice this problem early in the process so you know to work on it later. 

# 11.2.2 Cleaning the Dataset 

Transforming your dataset before visualizing it and fitting models is critical. It's easier to write code when variable names are concise and informative. Many functions in R, especially those in the `ggplot2` package, work best when datsets are in a "tidy" format. It's easier to do an analysis when you have only the information you need and very little extraneous data that can confuse your thought process. 

Since we're demonstrating one example of a cleaning process, we've worked through the whole thing and can preview the steps we took: 

1. Fix the variable names in the 2016 data 
4. Combine the datasets 
2. Pick variables 
1. Filter for the desired categories
3. Rename the variables 
3. Standardized the state names
5. Gather columns
3. Convert classes
3. Explore NAs

But in real life, data scientists often don't know the steps they'll need to take to prepare a dataset until they dive into the work. Learning what data transformations are needed requires exploration, trial and error, and clarity on the analytic questions you want to answer. 

After a lot of exploring, these are the steps we settled on for this particular analysis.  When you do your own, you will find all sorts of different things to transform. As you do more and more data analysis, your instincts for what to transform will also improve. 

## Fix the variable names in the 2016 data

When we print the 2016 dataset, we notice that the variable names are an actual year and an actual state name. Instead, we want them to be the actual variable names `Year` and `State Name`. To see that, we need to find which list element contains the 2016 data. The order of the list elements was set all the way back when we fed `map` our list of filenames. If we look at `filenames` again, we see that the 2016 dataset was stored in the fifth element: 

```{r}
filenames
```

Once know the 2016 dataset is stored in the fifth element of our list, we can pluck it out by using double brackets: 

```{r}
all_files[[5]]
```

We used `skip = 4` when we read in those datasets. That worked for all datasets except the fifth one. In that one, skipping four lines left out the variable name row. To fix it, we'll read the 2016 dataset again using the `read_csv` and the fifth element of `filenames`. We'll assign the newly read dataset to the fifth element of the `all_files` list: 

```{r}
all_files[[5]] <- read_csv(filenames[[5]], skip = 3)
```

Try printing `all_files` now. You can confirm we fixed the problem by verifying that all variable names are correct. 

## Select variables 

Now that we know all our datasets have variable names, we can start simplifying our datasets by picking the variables. This is a good place to think carefully about which variables we pick. This usually requires a fair amount of trial and error, but here is what we found we needed: 

 - Our analytic questions are about gender so let's pick the gender variables 
 - Later, we'll need to filter our dataset by disability category and program location so we'll want `SEA Education Environment` and `SEA Disability Category`
 - We want to make comparisons by state and reporting year so we'll also pick `State Name` and `Year`

Combining `select_at` and `contains` is a convenient way to pick these variables without writing a lot of code. Knowing that we want variables that contain the acronym "SEA" and variables that contain "male" in their names, we can pass those characters to `contains`:

```{r}
all_files[[1]] %>% 
    select_at(vars(Year, contains("State", ignore.case = FALSE), contains("SEA", ignore.case = FALSE), contains("male"))) 
```

That code chunk verifies that we got the variables we want, so now we will turn the code chunk into a function called `pick_vars`. We will then use `map` to feed our list of datasets, `all_files`, to the function. This will result in a newly transformed `all_files` list that contains six datasets, all with the desired variables. 

```{r}
pick_vars <- function(df) {
    df %>% 
        select_at(vars(Year, contains("State", ignore.case = FALSE), contains("SEA", ignore.case = FALSE), contains("male"))) 
}

all_files <- all_files %>% 
    map(pick_vars)
```

## Combine six datasets in a list into one dataset 

Now we'll turn our attention to combining the datasets in our list `all_files` into one dataset. We'll use `bind_rows`, which combines datasets by adding each one to the bottom of the one before it. The first step is to check and see if our datasets have the same number of variables and the same variable names. When we use `names` on our list of newly changed datasets, we see that each dataset's variable names are the same: 

```{r}
all_files %>% map(names)
```

That means that we can combine all six datasets into one using `bind_rows`. We'll call this newly combined dataset `child_counts`:

```{r}
child_counts <-  all_files %>% 
    bind_rows()
```

Since we know that a) each of our six datasets had eight variables and b) our combined dataset also has eight variables, we can conclude that all our rows combined together correctly. But let's use `str` to verify: 

```{r}
str(child_counts)
```

## Filter for the desired disabilities and age groups

We want to explore gender related variables, but our dataset has additional aggregate data for other subgroups. For example, we can use `count` to explore all the different disability groups in the dataset: 

```{r}
child_counts %>% 
    count(`SEA Disability Category`)
```

Since we will be visualizing and modeling gender variables for all students in the dataset, we'll filter out all subgoups except "All Disabilities" and the age totals: 

```{r}
child_counts <- child_counts %>% 
    filter(`SEA Disability Category` == "All Disabilities", 
           `SEA Education Environment` %in% c("Total, Age 3-5", "Total, Age 6-21")) 
```

## Rename the variables

In the next section we'll be preparing the dataset for visualization and modeling by "tidying" it. When we write code to transform datasets, we have to repeatedly type the column names so it's useful to change them to ones with reasonable lengths and without spaces. 

```{r}
child_counts <- child_counts %>% 
    rename(year = Year,
           state = "State Name", 
           age = "SEA Education Environment",
           disability = "SEA Disability Category", 
           f_3_5 = "Female Age 3 to 5", 
           m_3_5 = "Male Age 3 to 5", 
           f_6_21 = "Female Age 6 to 21", 
           m_6_21 = "Male Age 6 to 21")
```

## Clean state names

You might have noticed that some state names in our dataset are in upper case letters and some are in lower case letters: 

```{r}
child_counts %>% 
    count(state) %>% head()
```

If we leave it like this, R will treat the upper case states and lower case states of the same name as different states. We can use `mutate` and `tolower` to transform all the state names to lowercase letters. 

```{r}
child_counts <- child_counts %>% 
    mutate(state = tolower(state)) 
```

## Tidy the dataset 

Visualizing and modeling our data will be much easier if our dataset is in a "tidy" format. In his paper *Tidy Data*, Hadley Wickham defines tidy datasets where

>1. Each variable forms a column.
>2. Each observation forms a row.
>3. Each type of observational unit forms a table.

The gender variables in our dataset are spread across four columns, one representing a combination of gender and age range. We can use `gather` to bring the gender variable into one column. In this transformation, we create two new columns: a `gender` column and a `total` column. The `total` column will contain the number of students in each row's gender and age category.  

```{r}
child_counts <- child_counts %>% 
    gather(gender, total, f_3_5:m_6_21)
```

To make the values of the `gender` column more intuitive, we'll use `case_when` to transform the values to either "f" or "m":

```{r}
child_counts <- child_counts %>% 
    mutate(gender = case_when(
        gender == "f_3_5" ~ "f", 
        gender == "m_3_5" ~ "m", 
        gender == "f_6_21" ~ "f", 
        gender == "m_6_21" ~ "m", 
        TRUE ~ as.character(gender)
    ))
```

## Convert classes

The values in the `total` column represent the number of students from a specific year, state, gender, and age group. We know from the `chr` under their variable names that R is treating these values like characters instead of numbers. While R does a decent job of treating numbers like numbers when needed, it's much safer to prepare the dataset by changing these character columns to number columns. We'll use `dplyr::mutate_at` to conveniently change the count columns. 

```{r convert to numeric}
child_counts <- child_counts %>% 
    mutate(total = as.numeric(total)) 
child_counts
```

Changing these count columns from character classes to number classes resulted in two changes. First, the `chr` under these variable names has no changed to `dbl`, short for "double-precision". This lets us know that R recognizes these values as numbers with decimal points. Second, the blank values changed to `NA`. When R sees a character class value like `"4"`, it knows to change it to numeric class `4`. But there is no obvious number represented by a character class like `""`, so it changes it to `NA`: 

```{r}
# Convert a character to a number 
as.numeric("4")

# Convert a blank character to number
as.numeric("")
```

Similarly, the variable `year` needs to be changed from the character format to the date format. Doing so will make sure R treats this variable like a point in time when we plot our dataset. The package `lubridate` has a handy function called `ymd` that can help us. We just have to use the `truncated` argument to let R know we don't have a month and date to convert. 

```{r convert date}
library(lubridate)

child_counts <- child_counts %>% 
    mutate(year = ymd(year, truncated = 2))
```

## Explore and address NAs

You'll notice that some rows in the `total` column contain an `NA`. When we used `gather` to create a `gender` column, R created unique rows for every year, state, age, disability, and gender combination. Since the original dataset had both gender and age range store in a column like `Female Age 3 to 5`, R made rows that contained new no `total` value. For example, there is no student count for the `age` value "Total, Age 3-5" that also has the `gender` value for female students who were age 6-21. You can see that more clearly by sorting the dataset by year, state, and gender:

```{r}
child_counts %>% 
    arrange(year, state, gender)
```

We can simplify our dataset by removing these rows, leaving us with one row for each for 

 - females age 3-5
 - females age 6-21
 - males age 3-5 
 - males age 6-21 

... asscoiated with a state and reporting year: 

```{r remove NAs}
child_counts <- child_counts %>% 
    filter(!is.na(total)) 
```

We can verify that we have the four rows we want per year and state when we do our sort again: 

```{r verify rows}
child_counts %>% 
    arrange(year, state, gender)
```

# 11.3 How have Child Counts Changed Over Time? 

In the last section we focused on importing our dataset. In this section, we will shift our focus to exploring the dataset. We will use two tools for learning from this dataset. First, we'll use visualization to explore the number of students in special education over time. In particular, we'll compare the count of male and female students in time. Next, we'll use what we learn from our visualizations to quantify any differences that we see. 

# 11.3.2 Visualize the Dataset  

Plotting this many states can be overwhelming, so to start we'll make a subset of the dataset. We can use a function in the `dplyr` package called `top_n()` to help us learn which states have the highest average count of special education students: 

```{r}
child_counts %>% 
    group_by(state) %>% 
    summarise(mean_count = mean(total)) %>% 
    top_n(6, mean_count)
```

These six states have the highest mean count of students in special education over the six years we are examining. For reasons we will see in a later visualization, we are going to exclude outlying areas and freely associated states. That leaves us us with five states: California, Florida, New York, Pennsylvania, and Texas. We can get rid of all other states but these by using `filter()`. We'll call this new dataset `high_count`:

```{r filter most populated}
high_count <- child_counts %>% 
    filter(state %in% c("california", "florida", "new york", "pennsylvania", "texas"))
```

Now we can use `high_count` to do some initial exploration. Our analysis is about comparing counts of male and female students in special education, but with visualization is a great way to explore related curiosities. You may surprise yourself with what you find exploring by visualizing or summarising your datasets. You might come up with more interesting hypotheses, find that your initial hypothesis requires more data transformation, or find interesting subsets of the data--we'll saw a little of that in the surprisingly high `mean_count` of freely associated states in the `state` column). Let your curiosity and intuition drive this part of the analysis. It's one of the activities that makes data analysis a creative process. 

In that spirit, we'll start exploring by visualizing specific genders and age groups. Feel free to try these, but also try the other student groups for practice and more exploration. 
Try copying and running this code in your console to see what it does: 

```{r total female students over time}
high_count %>% 
    filter(gender == "f", age == "Total, Age 6-21") %>%
    ggplot(aes(x = year, y = total, color = state)) + 
    geom_freqpoly(stat = "identity") + 
    labs(title = "Count of female students in special education over time", 
         subtitle = "Ages 6-21")
```

That gives us a plot that has the years in the x-axis and a count of female students in the y-axis. Each line takes a different color based on the state it represents. 

Let's look at that closer: We used `filter` to subset our dataset for students who are female and ages 6 to 21. We used `aes` to connect visual elements of our plot to our data. We connected the x-axis to `year`, the y-axis to `total`, and the color of the line to `state`.

It's worth calling out one more thing, since it's a technique we'll be using as we explore further. Note here that, instead of storing our new dataset in a new variable, we filter the dataset then use the pipe operator `%>%` to feed it to `ggplot`. Since we're exploring freely, we don't need to create a lot of new variables we probably won't need later. This will change once we've learned a few new things about our dataset. At that point, we'll start targeting specific questions and having a newly named subsetted dataset will be convenient. 

We can also try the same plot, but subsetting for male students instead. We can use the same code we used for the last plot, but filter for the value "m" in the `gender` field: 

```{r total male students over time}
high_count %>% 
    filter(gender == "m", age == "Total, Age 6-21") %>%
    ggplot(aes(x = year, y = total, color = state)) + 
    geom_freqpoly(stat = "identity") + 
    labs(title = "Count of male students in special education over time", 
         subtitle = "Ages 6-21")
```

We've looked at each gender separately. What do these lines look like if we visualized the total amount of students each year per state? To do that, we'll need to add both gender values together and both age group values together. We'll do this using a very common combination of functions: `group_by` and `summarise`. 

```{r total students over time }
high_count %>% 
    group_by(year, state) %>% 
    summarise(n = sum(total)) %>% 
    ggplot(aes(x = year, y = n, color = state)) + 
    geom_freqpoly(stat = "identity")
```

So far we've looked at a few ways to count students over time. In each plot, we see that while counts have grown overall for all states, each state has different sized populations. Let's see if we can summarize that difference by looking at the median student count for each state over the years:

```{r median total per state}
high_count %>% 
    group_by(year, state) %>% 
    summarise(n = sum(total)) %>% 
    ggplot(aes(x = state, y = n)) + 
    geom_boxplot() + 
    labs(title = "Median student count", 
         subtitle = "All ages and genders")
```

The boxplots show us what we might have expected from our `freqpoly` plots before it. The highest median student count over time is California and the lowest is Pennsylvania. 

What have we learned about our data so far? The five states in the US with the highest total student counts (not including outlying areas and freely associated states) do not have similar counts to each other. The student counts for each state also appear to have grown over time. 

But how can we start comparing the male student count to the female student count? One way is to use a ratio: 

>In mathematics, a ratio is a relationship between two numbers indicating how many times the first number contains the second.
>Wikpedia

We can use the count of male students in each state and divide it by the count of each female student. The result is the number of times male students are in special education more or less than the female students in the same state and during the same year. Our coding strategy will be to 

 - Use `spread` to create separate columns for male and female students. 
 - Use `mutate` to create a new variable called `ratio`. The values in this column will be the result of dividing the count of male students by the count of female students 
 
Note here that the we can also accomplish this comparison by dividing the number of female students by the number of male students. In this case, the result would be the number of times female students are in special educatino more or less than male students. 

```{r male to female ratio over time }
high_count %>% 
    group_by(year, state, gender) %>% 
    summarise(total = sum(total)) %>%
    # Create new columns for male and female student counts
    spread(gender, total) %>% 
    # Create a new ratio column
    mutate(ratio = m / f) %>%
    ggplot(aes(x = year, y = ratio, color = state)) + 
    geom_freqpoly(stat = "identity") + 
    labs(title = "Male student to female student ratio over time", 
         subtitle = "Ages 3-21")
```

The result suggests that this ratio has declined between the years 2012 and 2017. In the next section, we'll explore ways to quantify the relationship between male student counts and female student counts. 

# 11.3.3 Model the Dataset 

At the start of this section, we chose to exclude outlying areas and freely associated states. You'll see in this visualization that there are some states that have a child count that is so high that it leaves a huge gap in x-axis values. This can be problematic when we try to learn from fitting our model later. 

```{r plot female students to male students}
child_counts %>% 
    filter(age == "Total, Age 6-21") %>%
    spread(gender, total) %>% 
ggplot(aes(x = f, y = m)) + 
    geom_point(size = 3, alpha = .5) + 
    geom_smooth() +
    labs(title = "Comparison of female students to male students in special education", 
         subtitle = "Counts of students in each state, ages 6-21", 
         x = "Female students", 
         y = "Male students", 
         caption = "Data: US Dept of Education"
    )
```

```{r where does the gap start}
child_counts %>% 
    filter(age == "Total, Age 6-21") %>%
    spread(gender, total) %>% 
    filter(f > 500000) %>%
    count(state)
```

```{r plot without outliers}
child_counts %>% 
    filter(age == "Total, Age 6-21") %>%
    spread(gender, total) %>% 
    filter(f < 500000) %>%
    ggplot(aes(x = f, y = m)) + 
    geom_point(size = 3, alpha = .5) + 
    labs(title = "Comparison of female students to male students in special education", 
         subtitle = "Counts of students in each state, ages 6-21.\nDoes not include outlying areas and freely associated states", 
         x = "Female students", 
         y = "Male students", 
         caption = "Data: US Dept of Education")
```

Model differences in identification by gender: 

```{r make model data}
model_data <- child_counts %>% 
    filter(age == "Total, Age 6-21") %>% 
    mutate(year = as.factor(year(year))) %>% 
    spread(gender, total) %>% 
    # Exclude outliers
    filter(f < 500000) %>% 
    # Compute male student to female student ratio 
    mutate(ratio = m / f) %>% 
    select(-c(age, disability))
```

```{r check factor sample sizes}
model_data %>% count(year)
```

```{r fit model for year predicting ratio }
ratio_year <- lm(ratio ~ year, data = model_data)
summary(ratio_year)
```

```{r average female to male ratio}
summary(model_data)
```

```{r visualize female to male ratio}
model_data %>% 
    gather(gender, students, c(f, m)) %>%
    ggplot(aes(x = year, y = students, color = gender)) + 
    geom_boxplot() + 
    labs(title = "Median male and female student counts in special education", 
         subtitle = "Ages 6-21. Does not include outlying areas and freely associated states", 
         x = "", 
         y = "", 
         caption = "Data: US Dept of Education")
```

# 11.4 Ask More Questions  

# 11.5 Aggregate Data as Context for Student Data 