# Walkthrough 5: Text analysis with social media data {#c11}

**Abstract**

This chapter explores tidying, transforming, visualizing, and analyzing text data. Data scientists in education are surrounded by text-based data sources like word processing documents, emails, and survey responses. Data scientists in education can expand their opportunities to learn about the student experience by adding text mining and natural language processing to their toolkit. Using Bluesky data, this chapter shows the reader practical tools for text analysis, including preparing text data, counting and visualizing words, and doing sentiment analysis. The chapter uses posts from #tidytuesday, a weekly data visualization challenge, to put these techniques in an education context. Data science tools in this chapter include transforming text into data frames, filtering datasets for keywords, running sentiment analysis and algorithms, and visualizing data.


## Topics emphasized

- Tidying data 
- Transforming data
- Visualizing data

## Functions introduced

- `sample_n()`
- `set.seed()`
- `tidytext::unnest_tokens()`
- `tidytext::get_sentiments()`
- `dplyr::inner_join()`
- `dplyr::anti_join()`

## Vocabulary 

- RDS files 
- Text analysis
- Stop words
- Tokenize 

## Chapter overview

The ability to work with many kinds of datasets is one of the great features of doing data science with programming. So far you've analyzed data in `.csv` files, but that's not the only way data is stored. Learning basic techniques for analyzing text gives you another way to understand the student experience.

In this chapter, you'll focus on analyzing textual data from Bluesky using the bskyr package [@bskyr]. We focus on this particular data source because it's relevant to a number of educational topics and questions. In addition, Bluesky data includes information about who posted, when they posted, the post's text, and a great deal more. This makes it especially well-suited for a group of text analysis techniques called natural language processing (often abbreviated NLP) [@hirschberg2015].

Note that we use #tidytuesday in this chapter and the next because it exemplifies learning-related data you might find in research. It's also straightforward to access from Bluesky. While this chapter dives deeply into the analysis of the text of posts, [Chapter 12](#c12) explores the nature of the interactions between individuals using the #tidytuesday hashtag.

### Background 

When thinking about data science in education, one tends to think about data stored in spreadsheets. But what can you learn about the student experience from text data? Take a moment to mentally review how much of your work day you spend either creating or consuming text data. In education, you're surrounded by it. Lessons are created in word processor documents, students submit assignments online, and the school community posts on public social media platforms. Text can be an authentic reflection of reality in schools, so how might we learn from it?

Even the most basic text analysis techniques will expand your data science toolkit. For example, you can use text analysis to count the number of key words that appear in open-ended survey responses. Or you can analyze word patterns in student responses or message board posts. 

Analyzing a collection of text is different from analyzing large numerical datasets because words don't have agreed-upon values. The number 2 will always be more than 1 and less than 3. The word "fantastic", on the other hand, has multiple meanings depending on interpretation and context. 

Using text analysis can help to broadly estimate meaning in the text. When paired with observations, interviews, and close review of the text, this approach can help educators learn from text data. In this chapter, you'll learn how to count the frequency of words in a dataset and associate those words with common feelings like positivity or joy. 

You'll learn these techniques using a dataset of posts from Bluesky. We encourage you to complete the walkthrough, then reflect on how what you've learned applies to other texts, like word processing documents or websites.  

### Data source

Social media platforms provide rich sources of text data that help you learn about communities. Since there's so much available, it's useful to narrow the posts so you can focus on your analytic questions. Hashtags are one way to do that. Here's an example:

>I'm trying to recreate some Stata code in R, anyone have a good resource for what certain functions in Stata are doing? #RStats #Stata

Bluesky recognizes any words that start with a "#" as a hashtag. The hashtags "#RStats" and "#Stata" make this post conveniently searchable. If you search for "#RStats", Bluesky returns all the posts containing that hashtag.

In this example, we'll be analyzing a dataset of posts that have the hashtag #tidytuesday. #tidytuesday is a community sparked by the work of one of the *Data Science in Education Using R* co-authors, Jesse Mostipak, who created the #r4ds community from which #tidytuesday was created. #tidytuesday is a weekly data visualization challenge. A great place to see examples from past #tidytuesday challenges is an interactive Shiny application (https://github.com/nsgrantham/tidytuesdayrocks).

The #tidytuesday hashtag returns posts about the weekly TidyTuesday practice, where folks learning R create and share data visualizations they made using tidyverse R packages.

### Methods

In this walkthrough, you'll be using simple methods -- counting words in a text dataset. Then you'll also use a more advanced technique called sentiment analysis to count and visualize words that have a positive association. Lastly, you'll learn how to get more context by selecting random rows of posts for closer reading or qualitative analysis.

## Load packages

For this analysis, you'll use the {tidyverse} and {here} packages. You'll also use the {tidytext} package for working with textual data [@R-tidytext]. If you haven't yet, you may need to install the {tidytext} package.

For instructions on installing packages, see the ["Packages" section](#c06p) of the ["Foundational Skills"](#c06) chapter.

Start by loading these packages:

```{r load packages, message = F, warning = F}
library(tidyverse)
library(tidytext)
```

## Import data

Next, load the data into your environment so you can start analyzing it. In [Chapter 12](#c12), you'll learn how to access this data through Bluesky's Application Programming Interface, or API.

You'll use the prepared dataset of #tidytuesday posts from Bluesky. Start by reading in this dataset:

```{r read data}
# raw_posts <- dataedu::bluesky_posts
raw_posts <- read_rds("posts.rds")
```

## View data

Return to your `raw_posts` dataset. Run `glimpse(raw_posts)` and notice the variables in this dataset. It's good practice to use functions like `glimpse()` or `str()` to look at the data type of each variable. For this walkthrough, you'll need to extract the text from the nested data structure and create an identifier for each post. 

## Process data

In this section you'll extract the text and transform the dataset so each row represents a word. After that, the dataset will be ready for exploring.

First, extract the text from the nested `record` structure. Each post's text is stored in `record$text`.

```{r pick columns}
posts <-
  raw_posts %>%
  mutate(post_id = row_number(),
         text = map_chr(record, ~pluck(.x, "text", .default = ""))) %>%
  filter(text != "") %>%
  select(post_id, text) %>%
  mutate(post_id = as.character(post_id))
```

Now the dataset has a column to identify each post and a column that shows the text that users posted. But each row has the entire post in the `text` variable, which makes it hard to analyze. If the dataset stays like this, you'd need to use functions on each row to count words. You can count words more efficiently if each row represents a single word. Splitting each sentence into individual words, one per row, is called "tokenizing". In their book *Text Mining With R*, @silge2017text describe tokens this way:

>A token is a meaningful unit of text, such as a word, that we are interested in using for analysis, and tokenization is the process of splitting text into tokens. This one-token-per-row structure is in contrast to the ways text is often stored in current analyses, perhaps as strings or in a document-term matrix.

Use `unnest_tokens()` from the {tidytext} package to transform the dataset of posts into a dataset of words.

```{r tokenize data}
tokens <-
  posts %>%
  unnest_tokens(output = word, input = text)

tokens
```

You use `output = word` to tell `unnest_tokens()` that you want the column of tokens to be called `word`. We use `input = text` to tell `unnest_tokens()` to tokenize the posts in the `text` column of the `posts` dataset. The result is a new dataset where each row has a single word in the `word` column and a unique ID in the `post_id` column that tells us which post the word appears in.

Notice that the `tokens` dataset has more rows than the `posts` dataset. This says a lot about how `unnest_tokens()` works. In the `posts` dataset, each row has an entire post and its unique ID. Since that unique ID is assigned to the entire post, each unique ID only appears once in the dataset. When you used `unnest_tokens()` to put each word on its own row, you broke each post into many words. This created additional rows in the dataset. And since each word in a single post shares the same ID for that post, an ID now appears multiple times in the new dataset. 

You're almost ready to start analyzing the dataset. There's one more step---removing common words that will make your analysis difficult. Words like "the" or "a" are in a category of words called "stop words". Stop words serve a function in verbal communication, but don't tell you much on their own. As a result, they clutter a dataset of useful words and make it harder to manage the volume of words during analysis. The {tidytext} package includes a dataset called `stop_words` that you'll use to remove rows containing stop words. You'll use `anti_join()` on the `tokens` dataset and the `stop_words` dataset to keep only rows that have words *not* appearing in the `stop_words` dataset. You'll also remove account names.

```{r remove stop words}
data(stop_words)

tokens <-
  tokens %>%
  anti_join(stop_words, by = "word") %>%
  filter(!str_detect(word, "\\.bsky\\.social$")) # removes account names from the words (tokens)
```

Why does this work? Let's look closer. `inner_join()` matches the observations in one dataset to another by a specified common variable. Any rows that don't have a match get dropped from the resulting dataset. `anti_join()` does the same thing as `inner_join()` except it drops matching rows and keeps the rows that don't match. This is convenient for the analysis because it removes rows from `tokens` that contain words in the `stop_words` dataset. When you call `anti_join()`, you're left with rows that don't match words in the `stop_words` dataset. These remaining words are the ones you'll be analyzing.

One final note before you start counting words: Remember when you first tokenized the dataset and passed `unnest_tokens()` the argument `output = word`? You conveniently chose `word` as the column name because it matches the column name `word` in the `stop_words` dataset. This makes the call to `anti_join()` simpler because `anti_join()` knows to look for the column named `word` in each dataset. 

## Analysis: counting words 

Now it's time to start exploring the newly cleaned dataset of posts. A good start is computing the frequency of each word and seeing which ones show up the most. You can pipe `tokens` to the `count()` function to do this: 

```{r count words}
tokens %>% 
    count(word, sort = TRUE) 
```

You pass `count()` the argument `sort = TRUE` to sort the `n` variable from the highest value to the lowest value. This makes it easy to see the most frequently occurring words at the top. Not surprisingly, "tidytuesday" was the most frequent word in this dataset. For future analyses, you may wish to remove it given that it was the basis of the search.

You may want to explore further by showing the frequency of words as a percent of the whole dataset. Calculating percentages like this is useful in education scenarios because it helps you make comparisons across different sized groups. For example, you can calculate the percentage of students in each classroom that receive special education services. 

In the posts dataset, you'll calculate the count of words as a percentage of all posts. Do that by using `mutate()` to add a column called `percent`. `percent` will divide `n` by `sum(n)`, which is the total number of words. Finally, multiply the result by 100:

```{r count words as percentage}
tokens %>%
  count(word, sort = TRUE) %>%
  # n as a percent of total words
  mutate(percent = n / sum(n) * 100)
```

Even at 622 appearances in our dataset, "tidytuesday" represents only about 6% of the total words. This makes sense when you consider our dataset contains 2,963 unique words. 

## Analysis: sentiment analysis

Now that you have a sense of the most frequently appearing words, it's time to explore sentiment using the posts dataset. Imagine you're an education consultant aiming to learn about the TidyTuesday data visualization ritual and its community. You know from the first part of the analysis that the token "dataviz" (short for data visualization) appeared frequently relative to other words. A good start would be to see how the appearance of that token in a post is associated with other positive words. 

You'll need to use a technique called "sentiment analysis" to get at the positivity of words in these posts. Sentiment analysis evaluates words for their emotional association. If you analyze words by the emotions they convey, you can start to explore patterns in large text datasets like the `tokens` data. 

Earlier you used `anti_join()` to remove stop words in the dataset. You're going to do something similar here to reduce the `tokens` dataset to words that have a positive association. You'll use a dataset called the "NRC Word-Emotion Association Lexicon" to identify words with a positive association. This dataset was published in a work called *Crowdsourcing a Word-Emotion Association Lexicon* [@mohammad2013].

You need to install a package called {textdata} to access the NRC Word-Emotion Association Lexicon dataset. Note that you only need to have the package installed. You do not need to load it with the `library(textdata)` command. 

If you don't already have it, install {textdata}: 

```{r install textdata, eval=FALSE}
install.packages("textdata")
```

To explore this dataset more, use a {tidytext} function called `get_sentiments()` to view words and their associated sentiment. If this is your first time using the NRC Word-Emotion Association Lexicon, you'll be prompted to download the NRC lexicon. Respond "yes" to the prompt and the NRC lexicon will download. Note that you'll only have to do this the first time you use the NRC lexicon. 

```{r read nrc}
#| include: false
nrc_sentiments <- read_rds("data/nrc_sentiments.Rds")
```

```{r view nrc}
#| eval: false
nrc_sentiments <- get_sentiments("nrc")
```

This returns a dataset with two columns. The first is `word` and contains a list of words. The second is the `sentiment` column, which contains an emotion associated with each word. This dataset is similar to the `stop_words` dataset. Note that this dataset also uses the column name `word`, which will again make it easy for us to match this dataset to the `tokens` dataset. 

### Count positive words

Begin working on reducing the `tokens` dataset down to only words that the NRC dataset associates with positivity. You'll start by creating a new dataset, `nrc_pos`, which contains the NRC words that have the positive sentiment. Then you'll match that new dataset to `tokens` using the `word` column that is common to both datasets. Finally, you'll use `count()` to total up the appearances of each positive word. 

```{r count positive words}
# Only positive in the NRC dataset
nrc_pos <-
    nrc_sentiments %>%
    filter(sentiment == "positive")

# Match to tokens
pos_tokens_count <-
  tokens %>%
  inner_join(nrc_pos, by = "word") %>%
  # Total appearance of positive words
  count(word, sort = TRUE) 

pos_tokens_count
```

You can visualize these words nicely by using {ggplot2} to show the positive words in a bar chart. There are 213 positive words total, which is hard to convey in a compact chart. Solve that problem by filtering the dataset to words that only appear 5 times or more. 

```{r fig11-1, results='hide', message = FALSE, fig.cap="Count of Words Associated with Positivity", fig.showtext = TRUE}
pos_tokens_count %>%
  filter(n >= 5) %>%
  ggplot(., aes(x = reorder(word, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(
    title = "Count of Words Associated with Positivity",
    subtitle = "Posts with the hashtag #tidytuesday",
    caption = "Data: Bluesky and NRC",
    x = "",
    y = "Count"
  ) +
  theme_minimal()
```

Note the use of `reorder()` when mapping the `word` variable to the x aesthetic. Using `reorder()` sorts the x-axis in descending order by the variable `n`. Sorting the bars from highest frequency to lowest makes it easier to identify and compare the most and least common words in the visualization. 

### "Dataviz" and other positive words

Earlier in the analysis we learned that "dataviz" was among the most frequently occurring words in this dataset. We can continue our exploration of TidyTuesday posts by seeing how many posts with "dataviz" also had at least one positive word from the NRC dataset. Looking at this can give us clues about how people in the TidyTuesday learning community view dataviz as a tool.

There are a few steps to this part of the analysis, so let's review a strategy. You'll need to use the `post_id` field in the `posts` dataset to filter the posts that contain the word `dataviz`. Then you'll use the `post_id` field in these `dataviz` posts to identify the ones that include at least one positive word.

How do you know which `post_id` values contain the word "dataviz" and which ones contain a positive word? Recall that the `tokens` dataset only has one word per row, which makes it easy to use functions like `filter()` and `inner_join()` to make two new datasets: one of `post_id` values that have "dataviz" in the `word` column and one of `post_id` values that have a positive word in the `word` column.

You'll explore the combinations of "dataviz" and any positive words in the `posts` dataset using these three ingredients: the `posts` dataset, a vector of `post_id`s for posts that have "dataviz" in them, and a vector of `post_id`s for posts that have positive words in them. Now that you have the strategy, it's time to write the code and see how it works.

First, you'll make a vector of `post_id`s for posts that have "dataviz" in them. You'll use this later to identify posts that contain "dataviz" in the text. Use `filter()` on the `tokens` dataset to keep only the rows that have "dataviz" in the `word` column. Name that new dataset `dv_tokens`. 

```{r post_id with dataviz}
dv_tokens <-
  tokens %>%
  filter(word == "dataviz")

dv_tokens
```

The result is a dataset that has `post_id` in one column and the word "dataviz" in the other column. Use `$` to extract a vector of `post_id` for posts that have "dataviz" in the text. This vector has hundreds of values, so you can use `head` to view just the first ten.

```{r just post_id}
# Extract post_id
head(dv_tokens$post_id)
```

Now you'll do this again, but this time, you'll make a vector of `post_id` for posts that have positive words in them. This will be used later to identify posts that contain a positive word in the text. We'll use `filter()` on the `tokens` dataset to keep only the rows that have any of the positive words in the `word` column. If you've been running all the code up to this point in the walkthrough, you'll notice that you already have a dataset of positive words called `nrc_pos`, which can be turned into a vector of positive words by typing `nrc_pos$word`. Use the `%in%` operator in the call to `filter()` to find only words that are in this vector of positive words. Name this new dataset `pos_tokens`.

```{r post_id with positive words}
pos_tokens <-
  tokens %>%
  filter(word %in% nrc_pos$word)

pos_tokens
```

The result is a dataset that has `post_id` in one column and a positive word from `tokens` in the other column. You'll again use `$` to extract a vector of `post_id` for these posts.

```{r just post_id 2}
# Extract post_id
head(pos_tokens$post_id)
```

That's a lot of `post_id`s, many of which are duplicates. You can make the vector of `post_id`s a little shorter. Use `distinct()` to get a data frame of `post_id`, where each `post_id` only appears once:

```{r using distinct}
pos_tokens <-
  pos_tokens %>%
  distinct(post_id)
```

Note that `distinct()` drops all variables except for `post_id`. For good measure, use `distinct()` on the `dv_tokens` data frame too:

```{r using distinct 2}
dv_tokens <-
  dv_tokens %>%
  distinct(post_id)
```

Now you have a data frame of `post_id` for posts containing "dataviz" and another for posts containing a positive word. Use these to transform the `posts` dataset. First filter `posts` for rows that have the "dataviz" `post_id`. Then create a new column called `positive` that will tell you if the `post_id` is from the vector of `post_id`s for positive words. Name this filtered dataset `dv_pos`.

```{r dataviz posts with positive variable}
dv_pos <-
  posts %>%
  # Only posts that have the dataviz post_id
  filter(post_id %in% dv_tokens$post_id) %>%
  # Is the post_id from our vector of positive words?
  mutate(positive = if_else(post_id %in% pos_tokens$post_id, 1, 0))
```

Take a moment to dissect how you use `if_else()` to create a `positive` column. You gave `if_else()` three arguments:

 - `post_id %in% pos_tokens$post_id`: a logical statement
 - `1`: the value of `positive` if the logical statement is true
 - `0`: the value of `positive` if the logical statement is false

The new `positive` column will take the value 1 if the `post_id` was in the `pos_tokens` dataset and the value 0 if the `post_id` was not in the `pos_tokens` dataset. Practically speaking, `positive` is 1 if the post has a positive word and 0 if it does not have a positive word.

And finally, look at what percent of posts that had "dataviz" in them also had at least one positive word:

```{r percent that had a positive word}
dv_pos %>%
  count(positive) %>%
  mutate(perc = n / sum(n))
```

About 41% of posts that have "dataviz" in them also have at least one positive word, and about 59% of them did not have at least one positive word. It's worth noting that this doesn't necessarily mean users had nothing positive to say in the other 59% of "dataviz" posts. You can't know why some posts had positive words and some didn't, but you can know that more dataviz posts had positive words than not. To put this in perspective, you might have a different impression if 5% or 95% of the posts had positive words. 

Since the point of exploratory data analysis is to explore and develop questions, let's continue to do that. In this last section we'll review a random selection of posts for context.

### Taking a close read of randomly selected posts

This dataset is large, so you started by zooming out to summarize the data. But it's also useful to zoom in and read some of the posts. This will help you build intuition and context about how users talk about TidyTuesday in general. Even though this doesn't lead to quantitative findings, it helps you learn more about the underlying post content.

Start by making a dataset of posts that has positive words from the NRC dataset. Earlier in this walkthrough, you made a dataset of posts that had "dataviz" and a column for positive words. Reuse that technique, but on the dataset of all posts instead of only the "dataviz" ones.

```{r positive posts}
pos_posts <-
  posts %>%
  mutate(positive = if_else(post_id %in% pos_tokens$post_id, 1, 0)) %>%
  filter(positive == 1)
```

Again, you're using `if_else` to make a new column called `positive` that takes its value based on whether `post_id %in% pos_tokens$post_id` is true or not.

You can use `slice()` to help you pick the rows. When you pass `slice()` a row number, it returns that row from the dataset. For example, you can select the 1st and 3rd row of our posts dataset this way:

```{r first and third post}
posts %>%
  slice(1, 3)
```

Randomly selecting rows from a dataset is a great technique to have in your toolkit. Random selection helps you avoid biases when picking rows to review.

Here's one way to do that using base R:

```{r}
sample(x = 1:10, size = 5)
```

Passing `sample()` a vector of numbers and the size of the sample you want returns a random selection from the vector. Try changing the value of `x` and `size` to see how this works.

{dplyr} has a version of this called `sample_n()` that you can use to randomly select rows in the posts dataset. Using `sample_n()` looks like this:

```{r randomly review}
set.seed(2026)

pos_posts %>%
  sample_n(., size = 10)
```

That returned ten randomly selected posts. Let's look a little closer at how that works. You used `sample_n()`, which returns randomly selected rows from the posts dataset. You also specified that `size = 10`, which means `sample_n()` will give you 10 randomly selected rows. 

A few lines before that, you used `set.seed(2026)`. This helps us ensure that, while `sample_n()` theoretically plucks 10 random numbers, the readers can run this code and get the same result you did. Using `set.seed(2026)` at the top of your code makes `sample_n()` pick the same ten rows every time. Try changing `2026` to another number and notice how `sample_n()` picks a different set of ten numbers, but repeatedly picks those numbers until you change the argument in `set.seed()`. 

## Conclusion

The purpose of this walkthrough is to share code with you so you can practice some basic text-analysis techniques. Now it's time to keep learning by adapting this code to text-based files in your work setting. Try reading some of these and doing a similar analysis:

 - News articles
 - Procedure manuals
 - Open-ended responses in surveys

There are also advanced text analysis techniques to explore. Consider trying topic modeling (https://www.tidytextmining.com/topicmodeling.html) or finding correlations between terms (https://www.tidytextmining.com/ngrams.html), both described in [@silge2017text].

Finally, keep reading if you'd like to explore this dataset more. You'll use it further in the next chapter [on social network analysis](#c12), where you explore how to collect your own Bluesky data and analyze the interactions between individuals in the #tidytuesday community.