# Walkthrough 1

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = F)
here::set_here(path = "~/documents/data-science-in-education")
```

---
title: A quick introduction and tutorial on a cool social network analysis model for influence
author: ''
date: '2018-08-02'
slug: social-network-analysis-model-for-influence
categories: [r]
tags: []
description: Influence models seem complex but are actually just regressions.
meta_img: /_media/images/msurbanstem-soc.png
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE
)
```

Social network is increasingly common in education ([Sweet, 2016](https://www.tandfonline.com/doi/abs/10.1080/00461520.2016.1208093)).

In this walkthrough I wanted to share a bit on the second type, or an **influence** model.

First, here is an equation. 

$$
{ y }_{ it2 }\quad =\quad \rho \sum _{ i'\quad =\quad 1 }^{ n }{ { w }_{ ii' }^{  } } { y }_{ i't1 }^{  }\quad +\quad { \gamma  }_{ it1 }^{  }\quad +\quad { \varepsilon  }_{ i }^{  }
$$

Some outcome measured at a *second* time point (i.e., after a school year) for individual $i$:

$$
{ y }_{ it2 }
$$

Is predicted by that individual's *first* time point measure:

$$
{ \gamma  }_{ it1 }^{  }
$$
A residual term:

$$
{ \varepsilon  }_{ i }^{  }
$$
And, importantly, an influence term:

$$
\sum _{ i'\quad =\quad 1 }^{ n }{ { w }_{ ii' }^{  } } { y }_{ i't1 }^{  }\quad 
$$

This behemoth needs more breaking down. $\rho$ is just the coefficient, like $\beta$ in a regression equation.

This part says something like, for individual $i$, find all of the other individuals they have some relationship with.

$$
\sum _{ i'\quad =\quad 1 }^{ n }
$$

From $i' = 1$, or the first individual, up to, for an individual with ten relationships, so $n = 10$ for example, to $i' = 10$, do *this*:

$$
{ { w }_{ ii' }^{  } } { y }_{ i't1 }^{  }
$$

Multiply the weight, $w$ of the relationship (i.e., how strong the relationship is or how many interactions there were) between $i$ and $i'$ by $i'$'s value on the outcome at time 1.

That's really the (key to) the whole thing. The idea is that the influence term "captures" how your interactions with someone, over some period of time (between the first and second time points) impact some outcome. This model accounts for an individual's initial report of the outcome, i.e., their time 1 prior value, so it is a model for *change* in some outcome.

It's a big, highly-confusing model, but one that is super cool, because you can quantify "the network effect." It's also fundamentally a regression. That's really it. All the work goes into calculating the influence term.

With Sarah Galey, I wrote a little code with an example. That code is below. In another post, I'd like to try to dive into this more, and maybe in another write about the other type of model I mentioned above, one for exploring who has relationships with whom (or a *selection* model), related to a project focused around the [#NGSSchat hashtag](https://twitter.com/search?q=%23ngsschat&src=typd).

```{r}
library(dplyr) # need to install with install.packages("dplyr") if not already installed (just need to do first time)

data1 <- data.frame(nominator = c(2, 1, 3, 1, 2, 6, 3, 5, 6, 4, 3, 4), 
                    nominee = c(1, 2, 2, 3, 3, 3, 4, 4, 4, 5, 6, 6), 
                    relate = c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1))

data2 <- data.frame(nominee = c(1, 2, 3, 4, 5, 6), 
                    yvar1 = c(2.4, 2.6, 1.1, -0.5, -3, -1))

data3 <- data.frame(nominator = c(1, 2, 3, 4, 5, 6),
                    yvar2 = c(2, 2, 1, -0.5, -2, -0.5))

# merge data1 and data2
# note: we want the nominee's indegree because this is who the nominator is being exposed to

data <- left_join(data1, data2, by = "nominee")
data$nominee <- as.character(data$nominee) # this makes merging later easier

# calculate indegree in tempdata and merge with data
tempdata <- data.frame(table(data$nominee))
names(tempdata) <- c("nominee", "indegree") # rename the column "nominee"
tempdata$nominee <- as.character(tempdata$nominee) # makes nominee a character data type, instead of a factor, which can cause problems
data <- left_join(data, tempdata, by = "nominee")

# Calculating exposure and an exposure term that uses indegree, exposure_plus
data$exposure <- data$relate * data$yvar1
data$exposure_plus <- data$exposure * (data$indegree + 1)

# Calculating mean exposure
mean_exposure <-
    data %>%
    group_by(nominator) %>%
    summarize(exposure_mean = mean(exposure))

mean_exposure_plus <-
    data %>%
    group_by(nominator) %>%
    summarize(exposure_plus_mean = mean(exposure_plus))

# need a final data set with mean_exposure, mean_exposure_plus, degree, yvar1, and yvar2 added

mean_exposure_terms <- dplyr::left_join(mean_exposure, mean_exposure_plus, by = "nominator")

names(data2) <- c("nominator", "yvar1") # rename nominee as nominator to merge these
final_data <- dplyr::left_join(mean_exposure_terms, data2, by = "nominator")
final_data <- dplyr::left_join(final_data, data3, by = "nominator") # data3 already has nominator, so no need to change

# regression (linear models)

model1 <- lm(yvar2 ~ yvar1 + exposure_mean, data = final_data)
summary(model1)

model2 <- lm(yvar2 ~ yvar1 + exposure_plus_mean, data = final_data)
summary(model2)
```